Characterizing narrative entities with embeddings : a case study of character relationships in Les Misérables

### 2022-10-12 article/paper1.tex
write paper

In the field of \emph{Digital Humanities}, \emph{Distant Reading} tools (REF) allow researchers to quickly gain knowledge on textual corpora without actually reading them. Purposes of these methods are various, but can be mainly categorized into two groups: in the first case, these methods are used in order to tag, classify, or summary large quantities of texts, in order to quickly structure information or to deliver a speech over the whole studied corpus (REF). Methods in this case rely heavily on Big Data and make an extensive use of Machine Learning, often with the help of supervised methods. In the second case, researchers use computational methods to underline hidden structures in a small corpus or even a lone text, which helps them to refine their understanding of this corpora and reinforce stated hypotheses (REF). Methods in this setting can also rely on Machine Learning, but must typically be build with more caution and attention to details: corpora are small, analyses are closer to the work, and methods must be more transparent in order to appropriately interpret results. The use of exploratory tools and unsupervised methods is also preferred in this context, as it is less desirable to base methods on information coming from large external corpora. The proposed method typically belong to this second group of methods, as it is unsupervised and work on a single document.

When a lone (or a few) textual narrative is analyzed, a particular effort is made into studying \emph{narrative entities} (characters, event, location, etc.) used by the author in her/his book. Researchers are often interested by characterizing them and see how they are articulated to one another in the story. Various computational tools can help them in this tasks, to name a few, Named Entity Recognition tools (REF), Automatic Character Networks Extraction (REF), Sentiment Analysis (REF), Topic Modeling (REF), Textometry (REF), and Word Embeddings (REF). All these methods have been used in order to explicitly show hidden structures constructed by the author in her/his work. It permits to find regularities and patterns, and can help to categorize particular narrative constructions, writing styles, or genres. These kind of methods can be great complement to classical analyses of literacy works as they allow to efficiently summarized information which is otherwise quite diffuse.

@@

The idea behind this methodology comes from the field of automatic extraction and analysis of \emph{character networks} from literacy works (REF). When building character networks from a textual narrative, the most widespread method consists in dividing the studied work into $n$ \emph{textual units} $u_1, \ldots, u_n$, which can be, e.g., sentences, paragraphs, or chapters, and then counting characters co-occurrences in these units (REF). Usually, the text constituting these units is discarded and the resulting network displays edges which roughly represent an aggregated number of interactions between characters. However, by doing so, the aggregation occurs on various type of interactions and will give little information about the type of relationship which exist between characters. Various improvements where proposed in order to weight or sign (or both) the edges in the character networks, such as using Sentiment Analysis (REF) Topic Modeling (REF), or BLABLA (REF). A particular inspiration for the current work is the article by (REF), where authors also analyzed characters in \emph{Les Misérables} by building various signed and weighted networks, with the help of Sentiment Analysis and Topic Modeling. The current framework was build by expanding this idea of refining character relationships while keeping exploring directions as wide as possible. Embeddings appeared to us to be the proper tool for achieving this. As a matter of fact, with embeddings, the textual contents of units are transformed into workable mathematical objects (the vectors), usable for various tasks, while conserving a maximum of information. The framework has been further generalized in order to be applicable on various type of narrative entities, but the presented case remains the study of character relationships in \emph{Les Misérables}.

@@

Various methods can be performed on the contingency table $\mathbf{N}$ in order to extract information from it, such as Topic Modeling (REF), Sentiment Analysis (REF), or Textometry (REF) methods. Here, we make to choice to extract a lower dimensional, numeric representation of each units, in other words, an \emph{embedding of textual units}. \\
In section \ref{entity_embeddings}, these vectors of textual units are used as anchor points in order to also embed entities into the same space. Therefore, it is crucial that an interpretation about the directions or the regions of this embedding is possible, in order to properly interpret the localization of entity vectors (the relative position of entity vectors among themselves is generally insufficient). For that reason, we focus on embeddings of textual units which also contain \emph{vectors of words}: by examining the positions of entities relatively to word vectors, entities can be characterized. We propose two embeddings verifying this condition: Section \ref{ca_method} describes \emph{Correspondence Analysis (\textbf{CA})} and section \ref{wv_method} focuses on \emph{Pre-trained Word Vectors (\textbf{WV})}.

@@

Using \emph{Correspondence Analysis} (\textbf{CA}) in order to analyze textual resources has a long tradition (REF). It has the advantage to naturally provide an embedding space, the factorial map, where units are placed alongside word vectors, and allows the interpretation of the placement of units in terms of word frequency profiles. Units and words vectors in the embedding space have a direct interpretation in terms of chi2 distance between profiles.

@@

\emph{Pre-trained Word Vectors}$  $ (\textbf{WV}), based on methods such as Word2Vec (REF), Glove (REF), or fastText (REF), have recently received great attention from various fields (REFS). They are generally obtained through a training on a very large corpora, such as Wikipedia (REF) or Common Crawl (REF), and the resulting embedding contains a large quantity of word vectors. As shown by multiple studies (REF), these vectors are placed in order to reflect semantic and syntactic similarities between words, and are used in various applications (REF).

There exists multiple methods in order to find vectors for a \emph{group of words}, such as sentences (REF), paragraphs (REF), or documents (REF), based on pre-trained word vectors. These derived vectors are often used to apply a classification or clustering algorithm  (REF) on the newly embedded objects, or to query information (REF). In order to derive vectors for groups of words, the majority of methods use the frequencies of words in the objects, i.e. a table similar to $\mathbf{N}$, but apply various weighting scheme and normalization in order to reduce the effects of frequent words and to standardize vectors. In the present article, we use a recent methodology proposed in (REF) as it is compatible with multiple unit sizes and gives state-of-the-art results in many tasks.

@@

Characters were detected using \emph{Flair}\footnote{\url{https://github.com/flairNLP/flair}.} (REF) NER tools. In order to unify character and to further refine the detected characters list, we used hand-made lists of character and aliases from NER results. It resulted in the detection of 54 characters. The entities considered in table $\mathbf{E}$ (Figure \ref{occ_table}) are single character (54 entities) and character pairs (547 entities), resulting in a table of size $365 \times 601$. A character or a character-pairs presence is considered present iff characters are detected at least 2 times in the chapter.

@@

The \textbf{CA} method is fully automated, and results in vectors in a $364$-dimensional space, while the $\textbf{WV}$ methods is based on pre-trained word vectors using \emph{fastText} (REF) trained on Common Crawl.\footnote{\url{ https://fasttext.cc/docs/en/crawl-vectors.html}.} For French, the number of word-vectors $m$ is 2 millions and the dimension of the vector space is $300$.

@@

Here again, we see that associated words for the \textbf{WV} give a general tone to tomes and entities, while \textbf{CA} results are more specific and related to particular events which occurred for characters. As expected, volume constants give a quick summary of each volumes, especially with the \textbf{CA-REG} method (e.g. $V_2$ for the Battle of Waterloo, $V_4$ for the barricade event). Associated words with entities also seems accurate in describing them. Note that Cosette was not detected in volume 3 because she is not explicitly cited (she is often refereed as "the daugther of M. Leblanc"), and this also explains the absence of the Cosette-Valjean pair.

@@

Finally, the biggest weakness of this framework yet is the difficulty to validate its pertinence. Several other case studies, with results carefully scrutinized regarding prior knowledge, should be undertaken in order to see if it result are solid, but this type of experiments are expansive both in time and explanation length. Another idea could be to use an annotated corpora such as in (REF), where human annotators already classified character relationships in various categories, and to see if the presented method can actually retrieve the different categories. Such an experiment can be promising, but it currently need highly performative automatic tools in order to detect and unify character entities occurrences, which is currently missing. Nevertheless, first case studies gave promising results for this framework, and its flexibility could lead to various applications.

@@

Note that various other quantities of interest can also be computed in CA, such as 
\begin{align*}
p_\alpha &:= \frac{\lambda_\alpha}{\lambda_\bullet} :\text{\small the proportion of inertia expressed in $\alpha$,} \\
c^u_{i \alpha} &:= \frac{f_i x_{i\alpha}^2}{\lambda_\alpha} : \text{\small the contribution of unit $i$ to axis $\alpha$,} \\
c^w_{j \alpha} &:= \frac{g_j w_{j\alpha}^2}{\lambda_\alpha} : \text{\small the contribution of word $j$ to axis $\alpha$,} \\
h^u_{i \alpha} &:= \frac{x_{i\alpha}^2}{\sum_{\alpha} x_{i\alpha}^2} : \text{\small the contribution of  axis $\alpha$ to unit $i$,} \\
h^w_{j \alpha} &:= \frac{w_{j\alpha}^2}{\sum_{\alpha} w_{j\alpha}^2} : \text{\small the contribution of  axis $\alpha$ to word $j$,}
\end{align*}
For a detailed interpretation of these different quantities, see (REF).

@@

This methods is justified and detailed in (REF). Let $\mathbf{w}_1, \ldots, \mathbf{w}_v$ be pre-trained word vectors which appear in the studied corpus, and the $(n \times v)$ table $\mathbf{N}$ counting the frequency of these words in the $n$ textual units. We first construct the \emph{uncentered vectors} $\widetilde{\mathbf{x}}_i$ of each unit $i$ with
\begin{equation}
\widetilde{\mathbf{x}}_i = \sum_{j = 1}^v \frac{n_{ij}}{n_{i \bullet}} \frac{a}{a + \frac{n_{\bullet j}}{n_{\bullet \bullet}}} \mathbf{w}_j,
\end{equation}
where $a > 0$ is an hyperparameter which gives less importance to frequent words as $a \to 0$. In this article, we set $a$ to the recommended value of $0.01$. Let $\widetilde{\mathbf{X}}$ be the matrix whose columns are vectors $\widetilde{\mathbf{x}}_i$, and $\mathbf{u}$ be its first singular vector. We compute \emph{vectors} $\mathbf{x}_i$ of each units $i$ with
\begin{equation}
\mathbf{x}_i = \widetilde{\mathbf{x}}_i - \mathbf{u}\mathbf{u}^\top \widetilde{\mathbf{x}}_i.
\end{equation}
This last equation act like a \emph{centration} of unit vectors in the direction of the first singular vector $\mathbf{u}$.

%% Define the bibliography file to be used
\bibliography{charnet}

@@

%%
%% End of file

>>>>

In the field of \emph{Digital Humanities}, \emph{Distant Reading} tools \cite{Moretti2014} allow researchers to quickly gain knowledge on textual corpora without actually reading them. Purposes of these methods are various, but can be mainly categorized into two groups: in the first case, these methods are used in order to tag, classify, or summarize large quantities of texts, in order to quickly structure information or to deliver a speech over the whole studied corpus \cite{Underwood2017}. Methods in this case rely heavily on Big Data and make an extensive use of Machine Learning, often with the help of supervised methods. In the second case, researchers use computational methods to underline hidden structures in a small corpus or even a lone text, which helps them to refine their understanding of this corpora and reinforce stated hypotheses \cite{Eve2017}. Methods in this setting can also rely on Machine Learning, but must typically be build with more caution and attention to details: corpora are small, analyses are closer to the work, and methods must be more transparent in order to appropriately interpret results. The use of exploratory tools and unsupervised methods is also preferred in this context, as it is less desirable to base methods on information coming from large external corpora. The proposed method typically belong to this second group of methods, as it is unsupervised and work on a single document.

When a lone (or a few) textual narrative is analyzed, a particular effort is made into studying \emph{narrative entities} (characters, event, location, etc.) used by the author in her/his book \cite{Schmid2010}. Researchers are often interested by characterizing them and see how they are articulated to one another in the story. Various computational tools can help them in this tasks, to name a few, Named Entity Recognition tools \cite{Agarwal2013, Chaturvedi2017, Li2022}, Automatic Character Networks Extraction \cite{Labatut2019}, Sentiment Analysis and Topic Modeling \cite{Min2019}, Textometry \cite{Novakova2019}, and Word Embeddings \cite{Grayson2016, Heuser2017, Kerr2017}. All these methods have been used in order to explicitly show hidden structures constructed by the author in her/his work. It permits to find regularities and patterns, and can help to categorize particular narrative constructions, writing styles, or genres. These kind of methods can be great complement to classical analyses of literacy works as they allow to efficiently summarized information which is otherwise quite diffuse.

@@

The idea behind this methodology comes from the field of automatic extraction and analysis of \emph{character networks} from literacy works (see \cite{Labatut2019} for a survey). When building character networks from a textual narrative, one of the most widespread methods consists in dividing the studied work into $n$ \emph{narrative units} or \emph{contexts} $u_1, \ldots, u_n$, which can be, e.g., sentences, paragraphs, or chapters, and then counting the number of units where characters co-occurred \cite{Elsner2012, Lee2012, Rochat2014, Grener2017, Min2019}. Usually, the text constituting these units is discarded and the resulting network displays edges which roughly represent an aggregated number of interactions between characters. However, by doing so, the aggregation occurs on various type of interactions and will give little information about the type of relationship which exist between characters. Various improvements where proposed in order to weight \cite{Sack2014} or sign \cite{Krishnan2015} (or both \cite{Min2019}) the edges in the character networks. A particular inspiration for the current work is the article by Min and Park \cite{Min2019}, where authors also analyzed characters in \emph{Les Misérables} by building various signed and weighted networks, with the help of Sentiment Analysis and Topic Modeling. The current framework was build by expanding this idea of refining character relationships while keeping exploring directions as wide as possible. Embeddings \cite{Incitti2023} appeared to us to be the proper tool for achieving this. As a matter of fact, with embeddings, the textual contents of units are transformed into workable mathematical objects (the vectors), usable for various tasks, while conserving a maximum of information. The framework has been further generalized in order to be applicable on various type of narrative entities, but the presented case remains the study of character relationships in \emph{Les Misérables}.

@@

Various methods can be performed on the contingency table $\mathbf{N}$ in order to extract information from it. Here, we make to choice to extract a lower dimensional, numeric representation of each units, in other words, an \emph{embedding of textual units}. \\
In section \ref{entity_embeddings}, these vectors of textual units are used as anchor points in order to also embed entities into the same space. Therefore, it is crucial that an interpretation about the directions or the regions of this embedding is possible, in order to properly interpret the localization of entity vectors (the relative position of entity vectors among themselves is generally insufficient). For that reason, we focus on embeddings of textual units which also contain \emph{vectors of words}: by examining the positions of entities relatively to word vectors, entities can be characterized. We propose two embeddings verifying this condition: Section \ref{ca_method} describes \emph{Correspondence Analysis (\textbf{CA})} and section \ref{wv_method} focuses on \emph{Pre-trained Word Vectors (\textbf{WV})}.

@@

Using \emph{Correspondence Analysis} (\textbf{CA}) in order to analyze textual resources has a long tradition \cite{Lebart2019}. It has the advantage to naturally provide an embedding space, the factorial map, where units are placed alongside word vectors, and allows the interpretation of the placement of units in terms of word frequency profiles. Units and words vectors in the embedding space have a direct interpretation in terms of chi2 distance between profiles.

@@

\emph{Pre-trained Word Vectors}$  $ (\textbf{WV}), based on methods such as Word2Vec \cite{Mikolov2013}, GloVe \cite{Pennington2014}, fastText \cite{Bojanowski2017}, or Bert \cite{Devlin2018} have received great attention from various fields in the last decade. They are generally obtained through a training on a very large corpora, such as Wikipedia or Common Crawl, and the resulting embedding contains a large quantity of word vectors. As shown by multiple studies (see \cite{Li2017} for a survey), these vectors are placed in order to reflect semantic and syntactic similarities between words, and are used in various applications. We will focus here on \emph{static} word embeddings, where word vectors are fixed and do not depends on their context, like Word2Vec or fastText, because we need to have a interpretable regions of the embedding space.

There exists multiple methods in order to find vectors for a \emph{group of words}, such as sentences \cite{Grener2017, Arora2017}, paragraphs \cite{Le2014}, or documents \cite{Kusner2015}, based on pre-trained word vectors. These derived vectors are often used to apply a classification or clustering algorithm on the newly embedded objects, or to query information \cite{Kusner2015, Arora2017}. In order to derive vectors for groups of words, the majority of methods use the frequencies of words in the objects, i.e. a table similar to $\mathbf{N}$, but apply various weighting scheme and normalization in order to reduce the effects of frequent words and to standardize vectors. In the present article, we use a methodology proposed in \cite{Arora2017} as it is compatible with multiple unit sizes and gives state-of-the-art results in many tasks.

@@

Characters were detected using \emph{Flair}\footnote{\url{https://github.com/flairNLP/flair}.} NER tools \cite{Akbik2018}. In order to unify character and to further refine the detected characters list, we used hand-made lists of character and aliases from NER results. It resulted in the detection of 54 characters. The entities considered in table $\mathbf{E}$ (Figure \ref{occ_table}) are single character (54 entities) and character pairs (547 entities), resulting in a table of size $365 \times 601$. A character or a character-pairs presence is considered present iff characters are detected at least 2 times in the chapter.

@@

The \textbf{CA} method is fully automated, and results in vectors in a $364$-dimensional space, while the $\textbf{WV}$ methods is based on pre-trained word vectors using \emph{fastText} \cite{Bojanowski2017} trained on Common Crawl.\footnote{\url{ https://fasttext.cc/docs/en/crawl-vectors.html}.} For French, the number of word-vectors $m$ is 2 millions and the dimension of the vector space is $300$.

@@

Here again, we see that associated words for the \textbf{WV} give a general tone to tomes and entities, while \textbf{CA} results are more specific and related to particular events which occurred for characters. As expected, words associated with volume constants give a short overview of each volumes, especially with the \textbf{CA-REG} method (e.g. $V_2$ for the Battle of Waterloo, $V_4$ for the barricade event). Associated words with entities also seems accurate in describing them. Note that Cosette was not detected in volume 3 because she is not explicitly cited (she is often refereed as "the daugther of M. Leblanc"), and this also explains the absence of the Cosette-Valjean pair.

@@

Finally, the biggest weakness of this framework yet is the difficulty to validate its pertinence. Several other case studies, with results carefully scrutinized regarding prior knowledge, should be undertaken in order to see if it result are solid, but this type of experiments are expansive both in time and human resources. Another idea could be to use an annotated corpora such as in \cite{Massey2015}, where human annotators already classified character relationships in various categories. We could, e.g., see if the presented method can actually retrieve these categories by assigning relationship entities to the category-word with the highest similarity value. Such an experiment can be promising, but it requires an efficient automatic entity tagger, in order to detect and especially unify characters in a large quantity of documents, and unfortunately, this tool do not exist yet. Nevertheless, first case studies gave promising results for this framework, and its flexibility could lead to various applications.

@@

Note that various other quantities of interest can also be computed in CA, such as 
\begin{align*}
p_\alpha &:= \frac{\lambda_\alpha}{\lambda_\bullet} :\text{\small the proportion of inertia expressed in $\alpha$,} \\
c^u_{i \alpha} &:= \frac{f_i x_{i\alpha}^2}{\lambda_\alpha} : \text{\small the contribution of unit $i$ to axis $\alpha$,} \\
c^w_{j \alpha} &:= \frac{g_j w_{j\alpha}^2}{\lambda_\alpha} : \text{\small the contribution of word $j$ to axis $\alpha$,} \\
h^u_{i \alpha} &:= \frac{x_{i\alpha}^2}{\sum_{\alpha} x_{i\alpha}^2} : \text{\small the contribution of  axis $\alpha$ to unit $i$,} \\
h^w_{j \alpha} &:= \frac{w_{j\alpha}^2}{\sum_{\alpha} w_{j\alpha}^2} : \text{\small the contribution of  axis $\alpha$ to word $j$,}
\end{align*}
For a detailed interpretation of these different quantities, see \cite{Lebart2019}.

@@

This methods is justified and detailed in \cite{Arora2017}. Let $\mathbf{w}_1, \ldots, \mathbf{w}_v$ be pre-trained word vectors which appear in the studied corpus, and the $(n \times v)$ table $\mathbf{N}$ counting the frequency of these words in the $n$ textual units. We first construct the \emph{uncentered vectors} $\widetilde{\mathbf{x}}_i$ of each unit $i$ with
\begin{equation}
\widetilde{\mathbf{x}}_i = \sum_{j = 1}^v \frac{n_{ij}}{n_{i \bullet}} \frac{a}{a + \frac{n_{\bullet j}}{n_{\bullet \bullet}}} \mathbf{w}_j,
\end{equation}
where $a > 0$ is an hyperparameter which gives less importance to frequent words as $a \to 0$. In this article, we set $a$ to the recommended value of $0.01$. Let $\widetilde{\mathbf{X}}$ be the matrix whose columns are vectors $\widetilde{\mathbf{x}}_i$, and $\mathbf{u}$ be its first singular vector. We compute \emph{vectors} $\mathbf{x}_i$ of each units $i$ with
\begin{equation}
\mathbf{x}_i = \widetilde{\mathbf{x}}_i - \mathbf{u}\mathbf{u}^\top \widetilde{\mathbf{x}}_i.
\end{equation}
This last equation act like a \emph{centration} of unit vectors in the direction of the first singular vector $\mathbf{u}$.

%% Define the bibliography file to be used
\bibliographystyle{apalike}
\bibliography{charnet}

@@
